A Bidirected String Graph Model for Genome Assembly

### 2012-12-06 presentation/paper.tex
paper: compute-overlaps

\subsubsection{Algorithm}

@@

>>>>

\subsubsection{Algorithms}

@@

The naive way to do this is to consider every pair of reads in the input,
including only distinct pairs but also including the special cases of each read
paired with itself, and search for each of the types of overlaps between the two
reads in each pair.  If there are $N$ reads in the input, this algorithm will
need to consider $\frac{N^2}{2} + N$ pairs of reads, and the time spent on each
read pair will depend several factors, including the minimum overlap length and
the read lengths.  But even if we could somehow spend constant time on each read
pair, this algorithm is much too slow because the running time rises quadraticly
with respect to the number of reads.  This is important because there can easily
be millions of reads in the input.

A number of faster algorithms for computing overlaps have been designed.  One of
the easier ones to understand is the {\it seed-and-extend} algorithm.  It works
by indexing the reads by short subsequences of a certain length (less than or
equal to than the minimum overlap length) and then only comparing reads that
actually share a subsequence.  Each of these short subsequences is called a {\it
seed}.  Two reads may share a seed in either a forward or reverse-complement
manner.  Given two reads that share a seed, the seed can be extended outwards in
both directions in order to find a maximal matched region, that may be either
forward or reverse-complement, between the two reads.  Then, this maximal
matched region is compared against the definition of an overlap.  In particular,
it must be at least as long as the minimum overlap length, it must have least 2
extreme points over the two reads, and it cannot be the prefix or suffix of both
reads if the match does not extend to the opposite end of at least one of the
reads.  If all of these conditions are met and the same overlap was not already
found through another seed in the same two reads, this overlap is added to the
set of overlaps.

The running time of the seed-and-extend algorithm, as described above, is
typically much faster than the naive algorithm, although it is highly dependent
on the data.  For a string of the genome of length $G$ and with a seed length
$k$, there will be at most $G-k+1$ canonical seeds.\footnote{The canonical form
of a DNA sequence and its reverse-complement is the lexicographically lesser of
the two.  Here, we refer canonical form of each seed, because we wish to compare
reads that share a seed in either a forward or reverse-complement manner.}  For
each of these seeds, the reads containing it will be examined exhaustively in a
pairwise manner, which takes quadratic time with respect to the number of reads
sharing that seed.  Since the more densely reads are sampled from the genome,
the more reads will contain each seed,  the running time of the seed-and-extend
algorithm rises quadratically with regards to read density.  In addition, the
number of read pairs the algorithm considers for a specific seed also rises
quadratically with the number of times that seed is repeated in the genome due
to genomic repeats.  Overall, the seed-and-extend algorithm is not ideal and is
still a bottleneck of the entire assembly algorithm, despite the fact that it is
much faster than the naive algorithm (except in degenerate cases, such as a
``genome'' that is entirely a single repeated letter).

For this project, the seed-and-extend algorithm for computing overlaps was
implemented as the program {\tt compute-overlaps}.  It takes as input a file
that contains a list of reads, as well as a parameter that specifies the minimum
overlap length $L$.  It produces as output (via a data structure serialized to a
file) all exact pairwise overlaps of minimum length $L$ among reads in the input
file, including reverse-complement overlaps and self-overlaps.  The running time
of the algorithm was discussed above.  The memory usage is linear with respect
to the total length of the reads and the number of overlaps, which likely could
be improved.
